\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-26 chapters/conclusion.tex
Comments Marc

\subsubsection{Regularized spatiotemporal beamforming}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (xDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher accuracy in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\cite{VanDenKerchove2022}.

@@

@@

\subsubsection{Healthy control study}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
$26.38\pm3.15$ years.
This dataset gives us insight in the ERP dynamics in overt, covert and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA, as well as that the
amplitude effect observed for covert VSA also holds for split VSA.

@@

>>>>

\subsubsection{\ac{erp} decoders}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (XDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher accuracy in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\cite{VanDenKerchove2022}.

@@

\todo{tensor}
\todo{wcble}

@@

\subsubsection{Gaze-independent \ac{bci} healthy controls}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
$26.38\pm3.15$ years.
This dataset gives us insight in the ERP dynamics in overt, covert and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA, as well as that the
amplitude effect\todomvh{not the right word, maybe the effect of covert VSA on
ERP...} observed for covert VSA also holds for split VSA.

@@

\subsection{Patient case studies}

### 2024-09-26 chapters/covert_align.tex
Comments Marc

\cite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert VSA compared to overt VSA.
This latency variability contributes to reduced covert VSA decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert VSA
decoding, but did not actually verify this hypothesis directly.
Additionally, \cite{hardiansyah2020single} developed a classifier for
covert VSA ERPs, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial ERP component latency
features on covert VSA inference yet did not attempt to correct the amplitude
features for these latencies.
\todo{Should we already mention Frenzel here?}
\cite{Frenzel2011} introduced a similar protocol to our split VSA
setting.
They showed that it is possible to perform split VSA and that, in this
case, visuospatial attention and gaze direction can separately be decoded using
classical ERP techniques.
To the best of our
knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
However, \cite{Frenzel2011} considered their interface only for the case
where the user actively intends to select both targets determined by the gaze
and the VSA.
In the split VSA setting considered in our work, we rather instruct the participant to ignore
the distractor and only attend the cued target, since we are interested in
decoding the visuospatial attention only.

@@

\subsection{Data collection}
We evaluate our approach on a
publicly available dataset and one specifically recorded for this study, designed to
probe different modalities of covert VSA.

@@

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different VSA conditions (overt, covert and split),
illustrated in Figure~\ref{fig:interface}.
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The VSA conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

@@

Participants signed the informed consent form and were seated at a distance of
60 cm before a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \cite{Treder2010} (Figure~\ref{fig:layout_targets}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding and, as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, Figure~\ref{fig:layout_intense}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.

@@

@@

\subsection{Data preprocessing and analysis}
Analysis was performed using Python and the MNE software package (version
1.3.1)~\cite{Gramfort2013}.
All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
Bad channels in the data were automatically detected using the RANSAC
method~\cite{Fischler1981} and rejected.
The recorded EEG was re-referenced
offline to the average of the mastoid electrodes TP9 and TP10.
Next, the EEG signals were corrected for eye movement artifacts using an
Independent Component Analysis (ICA).
Since we have access to EOG data for the CVSA-ERP dataset components correlating
significantly with the EOG were rejected.
For the BNCI2014-009 dataset, ICA components were manually rejected.
Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
In both datasets, only 16 channels were kept for
analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
PO8).

To evaluate performance, 6-fold cross-validation without shuffling was performed for both
datasets.
At each fold, classifiers were trained on five target selection blocks (300
epochs) and tested on one block (60 epochs) without overlap for CVSA-ERP.
For each subject and run in the BNCI2014-009 dataset, classifiers
were trained on five symbol selections (480 epochs) and tested on one symbol
selection (96 epochs) without overlap at each fold.
A window ranging from 0ms to 600ms after stimulus onset was used for CBLE and WCBLE.
With epochs ranging from -100ms to 700ms relative to stimulus onset, this
allows for extracting latencies ranging from -100ms to +100ms.
\section{Results}

@@

Although it is not immediately clear if WCBLE actively corrects for higher P3 jitter
present in covert and split VSA compared to overt VSA,
we justify our approach similar to \cite{hardiansyah2020single} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points towards higher P3 jitter as the main
contributing factor.
%Here, the subject using the GeoSpell interface must discriminate between target and
%non-target trials partly based on the stimulus characteristics
%(i.e. the selection of characters in their spelling application), compared to
%discrimination based on spatial location such as is the case in the CVSA-ERP
While Figure~\ref{fig:erps} shows no visible smearing effect in
the shape of the ERPs, a more quantitative analysis on latencies in
Figure~\ref{fig:jitter} indicates the opposite.
To obtain comparable results across VSA conditions, WCBLE was
evaluated per session and trained on all combined VSA conditions as in
Section~\ref{sec:cross_results}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Two-sided Wilcoxon rank-sum tests on jitter, expressed as the inter-quartile
range (IQR) of the estimated latencies of target trials, revealed that overt
VSA exhibited significantly lower jitter than all other conditions for both datasets.
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.
Additionally \cite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert VSA by comparing performance between
overt and covert VSA, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert VSA decoding performance.
However, the large increase in performance for covert VSA in \cite{Arico2014} could also
be explained by overfitting on artifacts amplified by aligning, since their
method is not evaluated on unseen data, as opposed to ours.

@@

While \cite{hardiansyah2020single} decoded covert VSA more effectively by
contributing single-trial latency and amplitude features to decoder.
Contrary to our approach, they did not then correct these amplitude
features for the jitter in their latencies by e.g. aligning trials to achieve
better separability.
Hence, their approach would not in principle render the classifier more robust
to jitter.
Furthermore we incorporated estimated latency features in both CBLE and WCBLE,
yet only WCBLE improved covert VSA performance.
This shows that the incorporation of latency features is not the only driver of
covert VSA decoding performance increase.

>>>>

\textcite{Arico2014} observed higher variability in single-trial P3 peak
latencies relative to stimulus onset during covert VSA compared to overt VSA.
This latency variability contributes to reduced covert VSA decoding performance.
While they proposed an analysis and performance prediction method, they did not provide a
decoding solution.
They suggested that compensating for latency jitter could enhance covert VSA
decoding, but did not actually verify this hypothesis directly.
Additionally, \cite{Hardiansyah2020} developed a classifier for
covert VSA ERPs, exploiting single-trial latency features in combination with
amplitude features for classification with a support vector machine.
They demonstrated the positive influence of single-trial ERP component latency
features on covert VSA inference, yet did not attempt to correct the amplitude
features for these latencies.
\todo{Should we already mention Frenzel here?}
\textcite{Frenzel2011} introduced a similar protocol to our split VSA
setting.
They showed that it is possible to perform split VSA and that, in this
case, visuospatial attention and gaze direction can separately be decoded using
classical ERP techniques.
To the best of our
knowledge, this is the sole study that investigated split attention in ERP-based BCIs.
However, \textcite{Frenzel2011} considered their interface only for the case
where the user actively intends to select both targets determined by the gaze
and the VSA.
In the split VSA setting considered in our work, we rather instruct the participant to ignore
the distractor and only attend the cued target, since we are interested in
decoding the visuospatial attention only.
\todomvh{avoid overlap. it seems already explained in previous sentences. The reader can find details in
  [6].}

@@

\subsection{Data collection}
We evaluate our approach on the
publicly available BNCI2014-009 dataset~\cite{Aloise2012a} and one specifically recorded for this study, designed to
probe different modalities of covert VSA.
\todomvh{first explain BNCI2014-009 using fig. 6.1 or revert the announcement
in the previous sentence.}

@@

We recorded a dataset to validate our approach.
The Covert Visuospatial Attention ERP (CVSA-ERP) dataset
consists of 15 participants, mean age $26.34\pm3.04$ years.
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
Each subject performed different VSA conditions (overt, covert and split),
illustrated in Figure~\ref{fig:interface} (top row).
Using a hexagonal layout interface, similar to the visual Hex-o-Spell proposed
by \cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG, electrooculogram (EOG), and the
participant's eye gaze using eye tracking were recorded.
The VSA conditions described in the first row of Figure~\ref{fig:interface} are
considered.
\begin{figure*}
  \bigskip
	\begin{subfigure}{\linewidth}
		\caption{}
		\label{fig:interface}
		\includegraphics[width=\linewidth]{figures/covert_align/figure3a.pdf}
	\end{subfigure}

@@

Participants signed the informed consent form and were seated at a distance of
60 cm before a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(Figure~\ref{fig:layout_targets})\todo{figure interface in four panels and
update all refs below}.
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding and, as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, Figure~\ref{fig:layout_intense}) instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychopPy (version 2023.1.3)~\cite{Peirce2019}.

@@

\subsection{Data processing and analysis}
\subsubsection{Preprocessing}
Analysis was performed using Python and the MNE software package (version
1.3.1)~\cite{Gramfort2013}.
All datasets were band-pass filtered between 0.1Hz and 20Hz with a 4th-order Butterworth filter.
Bad channels in the data were automatically detected using the RANSAC
method~\cite{Fischler1981} and rejected.
The recorded EEG was re-referenced
offline to the average of the mastoid electrodes TP9 and TP10.
Next, the EEG signals were corrected for eye movement artifacts using an
Independent Component Analysis (ICA).
Since we have access to EOG data for the CVSA-ERP dataset components correlating
significantly with the EOG were rejected.
For the BNCI2014-009 dataset, ICA components were manually rejected.
Finally, the EEG signal is divided into epochs ranging from 100ms before stimulus onset to 700ms after stimulus onset and down-sampled to 128Hz.
In both datasets, only 16 channels were kept for
analysis (Fz, FCz, Cz, CPz, Pz, Oz, F3, F4, C3, C4, CP3, CP4, P3, P4, PO7 and
PO8).

@@

To evaluate performance, 6-fold cross-validation without shuffling was performed for both
datasets.
At each fold, classifiers were trained on five target selection blocks (300
epochs) and tested on one block (60 epochs) without overlap for CVSA-ERP.
For each subject and run in the BNCI2014-009 dataset, classifiers
were trained on five symbol selections (480 epochs) and tested on one symbol
selection (96 epochs) without overlap at each fold.
A window ranging from 0ms to 600ms after stimulus onset was used for CBLE and WCBLE.
With epochs ranging from -100ms to 700ms relative to stimulus onset, this
allows for extracting latencies ranging from -100ms to +100ms.

\section{Results}

@@

Although it is not immediately clear if WCBLE actively corrects for higher P3 jitter
present in covert and split VSA compared to overt VSA,
we justify our approach similar to \cite{Hardiansyah2020} by
observing that the increased discrimination performance of a machine-learning
model accounting for jitter forgoes the need for characterizing the underlying
physiological processes while still objectively quantifying the presence of
jitter between data classes.
Furthermore, some evidence points towards higher P3 jitter as the main
contributing factor.
%Here, the subject using the GeoSpell interface must discriminate between target and
%non-target trials partly based on the stimulus characteristics
%(i.e. the selection of characters in their spelling application), compared to
%discrimination based on spatial location such as is the case in the CVSA-ERP
While Figure~\ref{fig:erps} shows no visible smearing effect in
the shape of the ERPs, a more quantitative analysis on latencies in
Figure~\ref{fig:jitter} indicates the opposite.
To obtain comparable results across VSA conditions, WCBLE was
evaluated per session and trained on all combined VSA conditions as in
Section~\ref{sec:cross_results}.
Since all conditions have the P3 in common, the estimated latencies can be
interpreted as P3 latencies.
Two-sided Wilcoxon rank-sum tests on jitter, expressed as the inter-quartile
range (IQR) of the estimated latencies of target trials, revealed that overt
VSA exhibited significantly lower jitter than all other conditions for both datasets.
No other significant differences were found.
$p$-values were corrected for the family-wise error rate using Bonferroni
correction.
Additionally \cite{Arico2014} prove that P3 jitter does play a
non-negligible role in covert VSA by comparing performance between
overt and covert VSA, while including or excluding early VEPs.
This analysis showed that the absence of the N1 and other early VEPs is not the
only factor hampering covert VSA decoding performance.
However, the large increase in performance for covert VSA in \cite{Arico2014} could also
be explained by overfitting on artifacts amplified by aligning, since their
method is not evaluated on unseen data, as opposed to ours.

@@

\textcite{Hardiansyah2020} decoded covert \textcite{vsa} more effectively by
contributing single-trial latency and amplitude features to decode.
Contrary to our approach, they did not then correct these amplitude
features for the jitter in their latencies by e.g. aligning trials to achieve
better separability.
Hence, their approach would not in principle render the classifier more robust
to jitter.
Furthermore we incorporated estimated latency features in both CBLE and WCBLE,
yet only WCBLE improved covert VSA performance.
This shows that the incorporation of latency features is not the only driver of
covert VSA decoding performance increase.

### 2024-09-26 chapters/wcble.tex
Comments Marc

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli.
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform.
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological and clinical properties, that can be confounding factors in
the development of \acp{bci}.\todo{cite}
Latency is often assessed as the latency of \ac{erp} components in a subject
average or grand average.
Latencies of non-averaged, single-trial \ac{erp} waveforms could capture even more
relevant information, yet they are very hard to measure due to low \ac{snr} of
single trial \acp{erp}.
Within session variability of \ac{erp} latency, or \ac{erp} latency
\emph{jitter} has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the intended target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased ERP jitter.
\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
performance, and disproportionally so when the user is not directly gazing at
intended targets, such as is the case for for gaze-impaired patients.

@@

\subsubsection{Neural origins of latency jitter}
While the classic \ac{erp} model assumes a constant \ac{erp} response over trials,
research has shown that cortical evoked potentials show significant
trial-to-trial jitter~\cite{Truccolo2002}.
Experiments show that latency jitter in \acp{erp} is the result of the interacion of
the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
	Kisley1999, Curto2009, Arieli1996}.
The latency of an evoked potential is a combination of the modulation of oscillating activity that is
time-locked, but not necessarily phase-locked to stimulus onset, and a
delay caused by ongoing neurophysiological processes that are not necessarily related to the task.
\todo{Check this, \cite{Stokes2016,Mouraux2008}}
\todo{Read references and check if neural origins and
	generation model justify approach}

@@

The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
most prominent of which is the smearing effect as illustrated
in~\autoref{fig:smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
averaged per condition to cope with the unfavorable \ac{snr}?
When averaging over multiple identical signals with jittered latencies, the shape
and amplitude of the average do not reflect the properties
of the original signal.
Because the amplitude of the average is lower, the \ac{snr} will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting \ac{erp}
results and making inferences about amplitude or latency effects, because these
are entangled when using averaging~\cite{Woody1967, Pfefferbaum1980, McDowell2003, Verleger2005,
	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
	these references}.

@@

The most straight-forward method to estimate single-trial latency is
\emph{peak picking}
Here, the latency of an \ac{erp} is determined as the
time-point of it's maximum (or minimum) amplitude relative to stimulus
onset.
While straight-forward, this method does not perform well in low or negative
\ac{snr} conditions, unless combined with filtering.
Filtering can suppress the noise contaminating \ac{eeg} trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}.
Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{
  Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999,  Sparacino2002,
  Georgiadis2005,	D’Avanzo2011}
and in the time-frequency domain~\cite{
  Quiroga2003, Wang2007,Hu2010}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using \ac{pca}~\cite{Ouyang2017}, ICA~\cite{Townsend1999, Jung2001,
	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.

@@

The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation training phase.
  (1) The first-stage spatiotemporal binary classifier is trained on a set of
  epochs.
  (2) It is	then applied to time-shifted copies of the epochs to obtain scores
  and class probabilities	over time.
  (3) The medians of these probability distributions are assumed the class
  latencies.
  (4) The epochs are aligned to their corresponding	class latencies by
  shifting in time such that all latencies fall at the same moment.
  (5) The spatiotemporal classifier is then retrained on the aligned epochs
  for a	next iteration.
  (6) After the iterative process halts, the scores and	latencies obtained
  from the last iteration are used to train the second-stage
  classifier.
}{%
  fig:diagram-train
}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\fullpagefig{
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation test phase. (1) The
	first-stage	spatiotemporal binary classifier obtained from the training phase is
	applied to time-shifted copies of the epochs to obtain scores and class probabilities
	over time. (2) The medians of these probability distributions are assumed
	as the new class latencies. (3) The scores and class latencies are input to the
  trained second-stage classifier, which predicts the label of the epochs.
}{%
  fig:diagram-eval
}
\section{Algorithms}
\todo{Pr not italic}
\todo{re-insert comments}

>>>>

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli.
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform.\todomvh{refer to image in bci chapter}
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological and clinical properties, that can be confounding factors in
the development of \acp{bci}.\todo{cite}
Latency is often assessed as the timing of an \ac{erp} component peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to low \ac{snr} of
single trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased ERP jitter.
\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
performance, and disproportionally so when the user is not directly gazing at
intended targets, such as is the case for for gaze-impaired patients.

@@

\subsubsection{Neural origins of latency jitter}
While the classic \ac{erp} model assumes a constant \ac{erp} response over trials,
research has shown that cortical evoked potentials show significant
trial-to-trial jitter~\cite{Truccolo2002}.
Experiments show that latency jitter in \acp{erp} is the result of the interaction of
the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
	Kisley1999, Curto2009, Arieli1996}.
According to the Firefly model of event-related activity~\cite{Burgess2012},
the latency of an evoked potential results from a combination of the modulation
of oscillating activity that is time-locked, but not necessarily phase-locked to stimulus onset,
and a delay caused by ongoing neurophysiological processes that are not necessarily related to the task
\cite{Stokes2016,Mouraux2008}.
\todo{substantiate}

@@

The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
most prominent of which is the smearing effect as illustrated
in~\autoref{fig:smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
averaged per condition to cope with the unfavorable \ac{snr}?
When averaging over multiple identical responses with jittered latencies, the shape
and amplitude of the average do not reflect the properties
of the original signal.
Because the amplitude of the average is lower, the \ac{snr} will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting \ac{erp}
results and making inferences about amplitude or latency effects, because these
are entangled when using averaging~\cite{Woody1967, Pfefferbaum1980, McDowell2003, Verleger2005,
	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
	these references}.

@@

The most straight-forward method to estimate single-trial latency is
\emph{peak picking}
Here, the latency of an \ac{erp} is determined as the
time-point of it's maximum (or minimum) amplitude relative to stimulus
onset.
While straight-forward, this method does not perform well in low
\ac{snr} conditions, unless combined with filtering.
Filtering can suppress the noise contaminating \ac{eeg} trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}.
Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{
  Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999,  Sparacino2002,
  Georgiadis2005,	D’Avanzo2011}
and in the time-frequency domain~\cite{
  Quiroga2003, Wang2007,Hu2010}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using \ac{pca}~\cite{Ouyang2017}, ICA~\cite{Townsend1999, Jung2001,
	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.

@@

The W\ac{cble} algorithm is presented in Algorithm~\ref{alg:wcble}.
Its training phase is visualized in~Figure~\ref{fig:diagram-train}.
The initial training epochs $\{\mat{X}_n^{(1)}\}_{n=1}^N$ are set to $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N$.
At every iteration, classifier $\mathcal{C}$ is trained like in \ac{cble}:
\begin{equation}
  \theta^{(i)} =
  \mathrm{train}_\mathcal{C}(\{\mat{X}^{(i-1)}_n[:,s_1:s_2]\}_{n=1}^N,\mathbf{l}^\mathrm{train})
\end{equation}
Next, latency $s_{l_n}^{(i)}$ is determined for every epoch $\mat{X}^{(i)}$ corresponding
to its class label $l_n$ using equation~\ref{eq:prob-latency}.
Finally, the training epochs $\mat{X}^{(i+1)}$ for the next iteration are determined by aligning
each original training epoch to the latency $s_{l_n}^{(i)}$ corresponding to its respective class
label.
\begin{equation}
  \mat{X}^{(i+1)}_n = \mathrm{align}(\mat{X}_n^\mathrm{train}, s_{l_n}^{(i)}) \quad \forall \quad n=1,\ldots,N
\end{equation}
Aligning is performed by shifting and zero-padding the signal to the right if
the latency is negative relative to the time window onset, and to the left if
positive, by the difference between the latency and the window onset.
\fullpagefig{%
  \includegraphics[width=\textwidth]{figures/wcble/figure1.pdf}
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation training phase.
  (1) The first-stage spatiotemporal binary classifier is trained on a set of
  epochs.
  (2) It is	then applied to time-shifted copies of the epochs to obtain scores
  and class probabilities	over time.
  (3) The medians of these probability distributions are assumed the class
  latencies.
  (4) The epochs are aligned to their corresponding	class latencies by
  shifting in time such that all latencies fall at the same moment.
  (5) The spatiotemporal classifier is then retrained on the aligned epochs
  for a	next iteration.
  (6) After the iterative process halts, the scores and	latencies obtained
  from the last iteration are used to train the second-stage
  classifier.
}{%
  fig:diagram-train
}
The process halts after a fixed amount of iterations or when the estimated set
of latencies has been encountered before, indicating it ended up in a loop.
In the end, the procedure should result in enhanced classifier parameters $\theta^*$,
closer to those when there would be no jitter between epochs.
Note that using the median approach for robust latency estimation results in a
smoother yet longer convergence process compared to the maximum/minimum approach.
We can then apply the classifier with enhanced parameters $\theta^*$ in a \ac{cble}
manner to unseen epochs as illustrated in Figure~\ref{fig:diagram-eval} to obtain a
vector of scores over time as in~Section~\ref{sec:cble} and the estimated latencies as
in~Section~\ref{sec:robust-latency}.
\fullpagefig{
  \includegraphics[width=\textwidth]{figures/wcble/figure2.pdf}
}{%
  Schematic representation of the Woody Classifier Based Latency
  Estimation test phase. (1) The
	first-stage	spatiotemporal binary classifier obtained from the training phase is
	applied to time-shifted copies of the epochs to obtain scores and class probabilities
	over time. (2) The medians of these probability distributions are assumed
	as the new class latencies. (3) The scores and class latencies are input to the
  trained second-stage classifier, which predicts the label of the epochs.
}{%
  fig:diagram-eval
}
\todo{Pr not italic}
\todo{re-insert comments}

### 2024-09-26 figures/bci/paradigms.tikz.tex
Comments Marc

    \node[color=darkgray, align=center,font=\bfseries, anchor=north] at (0,{-(1+\margin)}) {passive};
    \node[color=darkgray, align=center,font=\bfseries, anchor=south] at (0,{1+\margin}) {active};
    \node[color=darkgray, align=center,font=\bfseries, anchor=east] at ({-(1+\margin)}, 0) {stimulus\\independent};
    \node[color=darkgray, align=center,font=\bfseries, anchor=west] at ({1+\margin},0) {stimulus\\dependent};

>>>>

    \node[color=darkgray, align=center,font=\bfseries, anchor=north] at
    (0,{-(1+\margin)}) {passive\\participation};
    \node[color=darkgray, align=center,font=\bfseries, anchor=south] at
    (0,{1+\margin}) {active\\participation};
    \node[color=darkgray, align=center,font=\bfseries, anchor=east] at ({-(1+\margin)}, 0) {stimulus\\independent};
    \node[color=darkgray, align=center,font=\bfseries, anchor=west] at ({1+\margin},0) {stimulus\\dependent};
